---
title: Getting familiar with R
author: ""
---

## Aim of this worksheet
  
After completing this worksheet, you should feel comfortable typing commands into the R console (or, [REPL](https://en.wikipedia.org/wiki/Read%E2%80%93eval%E2%80%93print_loop)) and into an [R Markdown](http://rmarkdown.rstudio.com/) document. In particular, you should know how to use values, variables, and functions, how to install and load packages, and how to use the built-in help for R and its packages.

## Values

R lets you store several different kinds of *values*. These values are the information that we actually want to do something with. 

One kind of value is a number. Notice that typing this number, either in an R Markdown document or at the console, produces an identical output

```{r}
42
```

Create a numeric value that has a decimal point:

```{r}

```

Of course numbers can be added together (with `+`), subtracted (with `-`), multiplied (with `*`), and divided (with `/`), along with other arithmetical operations. Let's add two numbers, which will produce a new number.

```{r}
2 + 2
```

Add two lines, one that multiplies two numbers, and another that subtracts two numbers.

Another important kind of value is a character vector. (Most other programming languages would call these strings.) These contain text. To create a string, include some characters in between quotation marks `""`. (Single quotation marks work too, but in general use double-quotation marks as a matter of style.) For instance:

```{r}
"Hello, beginning R programmer"
```

Create a string with a message to your instructor.

```{r}

```

Character vectors can't be added together with `+`. But they can be joined together with the `paste()` function.

```{r}
paste("Hello", "everybody")
```

Mimic the example above and paste three strings together.

```{r}

```

Now explain in a sentence what happened.

> 

Another very important kind of value are logical values. There are only two of them: `TRUE` and `FALSE`.

```{r}
# This is true
TRUE

# This is false
FALSE
```

Notice that in the block above, the `#` character starts a *comment*. That means that from that point on, R will ignore whatever is on that line until a new line begins.

Logical values aren't very exciting, but they are useful when we compare other values to one another. For instance, we can compare two numbers to one another.

```{r}
2 < 3
2 > 3
2 == 3
```

What do each of those comparison operators do? (Note the double equal sign: `==`.)

>

Create your own comparisons between numeric values. See if you can create a comparison between character vectors.

```{r}

```

## Variables

We wouldn't be able to get very far if we only used values. We also need a place to store them, a way of writing them to the computer's memory. We can do that by *assignment* to a variable. Assignment has three parts: it has the name of a variable (which cannot contain spaces), an assignment operator, and a value that will be assigned. Most programming languages use a rinky-dink `=` for assignment, which works in R too. But R is awesome because the assignment operator is `<-`, a lovely little arrow which tells you that the value goes into the variable. For example:

```{r}
number <- 42
```

Notice that nothing was printed as output when we did that. But now we can just type `a` and get the value which is stored in the variable.

```{r}
number
```

It works with character vectors too.

```{r}
computer_name <- "HAL 9000"
```

No output, but this works.

```{r}
computer_name
```

In the assignment above, what is the name of the variable? What is the assignment operator? What is the value assigned to the variable?

>

Notice that we can use variables any place that we used to use values. For example:

```{r}
x <- 2
y <- 5
x * y
x + 9
```

Explain in your own words what just happened.

> 

Now create two assignments. Assign a number to a variable and a character vector to a different variable.

```{r}

```

Now create a third variable with a numeric value and, using the variable with a numeric value from earlier, add them together.

```{r}

```

Can you predict what the result of running this code will be? (That is, what value is stored in `a`?)

```{r, eval=FALSE}
a <- 10
b <- 20
a <- b
a
```

Predict your answer, then run the code. What is the value stored in `a` by the end? Explain why you were right or wrong.

>

## Vectors

c

vector recycling

## Built-in functions

sum, mean, median

class
str
print


## Using the documentation

?help

## Data frames

load a package

$, [



## Writing your own function


## Plots

Plot, title


## Download and load a CSV

download.file

read.csv

read_csv
